---
title: "Data Manipulation"
author: "Joddy"
date: "2024-02-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R Open Lab 2: Reading, writing, and managing multiple datasets

# Part 2: Data manipulation

## Merging data with dplyr

dplyr allows multiple datasets to be "joined" (terminology from SQL):

  Function	    Objective

  left_join()	  Keep all observations from the origin table

  inner_join()	Excludes all unmatched rows

  full_join()   Merge two datasets. Keeps all observations

  right_join()	Keep all observations from the destination table
  
## Join visualization 
  
For a visual explanation of joining data, we will go over https://go.ncsu.edu/dplyr-joins.

I found these visual aids really helpful when I was learning SQL (and also R joins), so you can refer to them in the future as you're learning.

## Load tidyverse package for data manipulation
```{r}
library(dplyr)
```

## Generate some examples
```{r}
df_primary <- tribble(
  ~ID, ~y,
  "A", 5,
  "B", 5,
  "C", 8,
  "D", 0,
  "F", 9)
df_secondary <- tribble(
  ~ID, ~z,
  "A", 30,
  "B", 21,
  "C", 22,
  "D", 25,
  "E", 29)

# Left join
left_join(df_primary, df_secondary, by ='ID')

# Inner join
inner_join(df_primary, df_secondary, by ='ID')

# Full join
full_join(df_primary, df_secondary, by = 'ID')

# Right join
right_join(df_primary, df_secondary, by = 'ID')

left_join(df_secondary, df_primary, by="ID")
```

How could we rewrite the right join as a left join?

In practice, we will typically use left joins, because they are easier to understand. The primary dataset is generally the dataset we want to "add columns" to, and left join is more intuitive for that task.

## Joins with multiple keys
```{r}
df_primary_m <- tribble(
  ~ID, ~year, ~items,
  "A", 2015,3,
  "A", 2016,7,
  "A", 2017,6,
  "B", 2015,4,
  "B", 2016,8,
  "B", 2017,7,
  "C", 2015,4,
  "C", 2016,6,
  "C", 2017,6)
df_secondary_m <- tribble(
  ~ID, ~year, ~prices,
  "A", 2015,9,
  "A", 2016,8,
  "A", 2017,12,
  "B", 2015,13,
  "B", 2016,14,
  "B", 2017,6,
  "C", 2015,15,
  "C", 2016,15,
  "C", 2017,13)
left_join(df_primary_m, df_secondary_m, by = c('ID', 'year'))
```

## NYC flights example

```{r}
# Load the datasets
library(nycflights13)
```


The NYC flights data (nycflights13 package) has several related datasets to join:

* airlines records two pieces of data about each airline: its carrier code and its full name. You can identify an airline with its two letter carrier code, making carrier the primary key.

```{r}
airlines
```

* airports records data about each airport. You can identify each airport by its three letter airport code, making faa the primary key.

```{r}
airports
```

* planes records data about each plane. You can identify a plane by its tail number, making tailnum the primary key.

```{r}
planes
```

* weather records data about the weather at the origin airports. You can identify each observation by the combination of location and time, making origin and time_hour the compound primary key.

```{r}
weather
```
```{r}
flights
```

### Foreign keys
A foreign key is a variable (or set of variables) that corresponds to a primary key in another table. For example:

flights$tailnum is a foreign key that corresponds to the primary key planes$tailnum.
flights$carrier is a foreign key that corresponds to the primary key airlines$carrier.
flights$origin is a foreign key that corresponds to the primary key airports$faa.
flights$dest is a foreign key that corresponds to the primary key airports$faa.

## Exploring the relationships in flights data

We will come back to these for our exercises to learn how to use joins to explore relationships between the datasets.

## World Bank population example

```{r}
# We'll use this to pull data
library(wbstats)
```

### Import and clean the data
```{r}
# The SP.POP.TOTL indicator of total population
pop_wb <- wb_data("SP.POP.TOTL", start_date = 2000, end_date = 2022)
pop_wb

# The SP.DYN.TFRT.INof total fertility rate in rates per woman
tfrt_wb <- wb_data("SP.DYN.TFRT.IN", start_date = 2000, end_date = 2022)
tfrt_wb

# Clean up the World Bank data
pop <- pop_wb %>%
  filter(!is.na(SP.POP.TOTL)) %>%
  select(country, date, SP.POP.TOTL) %>%
  rename(population = SP.POP.TOTL)

tfrt <- tfrt_wb %>%
  filter(!is.na(SP.DYN.TFRT.IN)) %>%
  select(country, date, SP.DYN.TFRT.IN) %>%
  rename(fertility = SP.DYN.TFRT.IN)

# Select subset from 2010
pop_2010 <- pop %>%
  filter(date == 2010) %>%
  select(-date)

tfrt_2010 <- tfrt %>%
  filter(date == 2010) %>%
  select(-date)
```

### Inner Join 
```{r}
# Inner join for the year 2010 of the population and fertility datasets
inner_2010 <- inner_join(pop_2010, 
                         tfrt_2010, 
                         by = "country")
inner_2010
```

### Full join
```{r}
# Full join for 2010
full_2010 <- full_join(pop_2010, 
                       tfrt_2010, 
                       by = "country")
full_2010

# Examine the rows that are missing fertility or population
full_2010 %>%
  filter(is.na(fertility))

full_2010 %>%
  filter(is.na(population))
```

### Left join
```{r}
left_2010 <- left_join(pop_2010, 
                       tfrt_2010, 
                       by = "country")
left_2010
```

Why do we get the same number of rows as the full join?

### Right join
```{r}
right_2010 <- right_join(pop_2010, 
                         tfrt_2010, 
                         by = "country")
right_2010

# How could we rewrite this using a left join?
left_tfrt_2010 <- left_join(tfrt_2010, 
                         pop_2010, 
                         by = "country")
left_tfrt_2010
```

### Anti join

```{r}
# anti_join(x, y) returns observations of x with no match in y
count(anti_join(pop, tfrt, by = c("country", "date")))
```

In this case, we get the rows with population values but no fertility values

## Many-to-many relationships
By default, dplyr guards against many-to-many relationships in equality joins by throwing a warning. These occur when both of the following are true:
  
1. A row in x matches multiple rows in y.

2. A row in y matches multiple rows in x.

This is typically surprising, as most joins involve a relationship of one-to-one, one-to-many, or many-to-one, and is often the result of an improperly specified join. Many-to-many relationships are particularly problematic because they can result in a Cartesian explosion of the number of rows returned from the join.

```{r}
# many rows of population and fertility for each country (one for each year)
cartesian_explosion <- full_join(tfrt, pop, by="country")

# dplyr protected us from running this join. If we really want a many-to-many join, specify like this:
cartesian_explosion <- full_join(tfrt, pop, by="country", relationship = "many-to-many")
```

Remember from the source data: 4991 rows in pop, 4631 rows in tfrt

many_to_many_explosion has 106605 rows!

How did we get so many rows? The full join matched the possible combinations of the countries and years.

```{r}
pop %>% filter(country=="Italy") # 23 rows
tfrt %>% filter(country=="Italy") # 22 rows
```

Cartesian product of these two tables is 23 * 22 = 506

```{r}
cartesian_explosion %>% filter(country=="Italy")
```

Each row in pop matched all 22 rows in tfrt.

## Sources for this tutorial
- https://www.guru99.com/r-dplyr-tutorial.html
- https://jmsallan.netlify.app/blog/joining-relational-tables-in-dplyr/
- https://r4ds.hadley.nz/joins.html
- https://dplyr.tidyverse.org/reference/mutate-joins.html

go.ncsu.edu/dvs-eval
