---
title: "Importing and exporting data in R"
author: "Joddy"
date: "2024-02-14"
output:
  html_document:
    toc: true # look this up for Ming tomorrow (3/12)
    # BTM topic model (Latex)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R Open Lab 2: Reading, writing, and managing multiple datasets

# Part 1: Importing and exporting data in R

## Getting and setting your path, and using paths in R

### Get the working directory
```{r}
getwd()

# Set the working directory (Linux and Mac)
# setwd("/Users/jimarche/R/r open labs/r_open_lab-1-read_write_join/data")

# What if you are on Windows?
# setwd("C:/Users/jimarche/R/r open labs/r_open_lab-1-read_write_join/data")
```

## 1. Importing data

#### Text file
read.table() is the most basic importing function. You can specify tons of different arguments in this function.

```{r}
stocks <- read.table("data/sample-5.tsv", sep="\t", skip=26, skipNul=TRUE, header=TRUE)
stocks <- stocks[,-1] # Get rid of the first column, which is blank

library(readr)

stocks <- read_tsv("data/sample-5.tsv", skip=20, col_select=-1) # col_select lets us delete column 1 in the read function
```

### Comma separated variable (csv)
```{r}
library(readr)

birds <- read_csv("data/birds.csv")

birds <- read_csv("data/birds.csv", col_names=TRUE, skip=0)
```

### Excel (xls, xlsx)
```{r}
library(readxl)

ufo <- read_excel("data/ufo.xlsx")

# using range= to: take the first 200 rows from ufo.xlsx, dropping the last 3 columns 
ufo <- read_excel("data/ufo.xlsx", col_names=TRUE, range="B1:H201")

# Loading a sheet from an excel file with multiple sheets
ufo <- read_excel("data/ufo.xlsx", sheet="ufo")

# read_xls for older excel files (.xls)
participants <- read_xls("data/participants.xls")
```

### Other data formats using Haven
```{r}
library(haven)

#  **SAS**: `read_sas()` reads .sas7bdat and .sas7bcat files and `read_xpt()` reads SAS transport files (version 5 and version 8).
# `write_sas()` writes .sas7bdat files.
airline <- read_sas("data/airline.sas7bdat")

# **SPSS**: `read_sav()` reads .sav files and `read_por()` reads the older .por files. `write_sav()` writes .sav files.
bankloan <- read_sav("data/bankloan.sav")

# **Stata**: `read_dta()` reads .dta files (up to version 15).
# `write_dta()` writes .dta files (versions 8-15).
auto <- read_dta("data/auto.dta")
```

## 2. Writing datasets from data.frame or tibble

### Writing .txt and .csv files
Loading mtcars data
```{r}
# Write data to txt file: tab separated values
write.table(mtcars, file = "data/mtcars.txt", sep = "\t",
            row.names = TRUE, col.names = NA)

# Write data to csv files:  
# decimal point = "." and value separators = comma (",")
write.csv(mtcars, file = "data/mtcars.csv")
```

### Using the readr package (tidyverse)

```{r}
# Loading mtcars data
data(mtcars)
library(readr)

# Writing mtcars data to a tsv file
write_tsv(mtcars, file = "data/mtcars.txt")
# Writing mtcars data to a csv file
write_csv(mtcars, file = "data/mtcars.csv")
```

### Writing .xlsx files
```{r}
library(writexl)

df <- data.frame(
  name = c("UCLA", "Berkeley", "Jeroen"),
  founded = c(1919, 1868, 2030),
  website = xl_hyperlink(c("http://www.ucla.edu", "http://www.berkeley.edu", NA), "homepage") )
df$age <- xl_formula('=(YEAR(TODAY()) - INDIRECT("B" & ROW()))')
write_xlsx(df, 'data/universities.xlsx')
```

### Exporting R data files

```{r}
# Saving and restoring one single R object (save a single object to a file):
saveRDS(mtcars, "data/mtcars.rds")
# Restore it under a different name
my_data <- readRDS("data/mtcars.rds")
```

### Saving and restoring one or more R objects

```{r}
ufo1 <- ufo[1:100,]
ufo2 <- ufo[101:200,]

# Save multiple objects
save(ufo1, ufo2, file = "data/ufo_first200.RData")
# To load the data again
load("data/ufo_first200.RData")
```

## What other file types have you worked with? Let's try to import them into R!

